---
title: "Human apicome LCM proteomics analysis"
output: html_notebook
author: "Roy Novoselsky"
date: "April 2024"
---

```{r}
rm(list=ls()) 
```

# Dependencies
```{r warning=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(ggrepel)
library(ggpubr)
library(readxl)
library(viridis)
library(tidyverse)
library(data.table)
library(gridExtra)
library(patchwork)
library(grid)

directory = getwd()
root_dir =  paste0(unlist(strsplit(directory, '/'))[-length(unlist(strsplit(directory, '/')))], collapse = '/')
source(paste(root_dir,"/functions.R",sep=""))
```

# load data & metadata
```{r}
dir_output = paste(directory,"output",sep="/")

data = read.csv(paste(dir_output,"data.csv",sep="/"),row.names=1)
rownames(data) = data$gene
metadata_features = read.csv(paste(dir_output,"metadata.csv",sep="/"),row.names=1)

```

# plot number of proteins detected
```{r}
data$gene=NULL

proteins_thresh = 1000

number_of_proteins = apply(data,2,function(x) sum(!is.na(x)))

meta = data.frame(proteins=number_of_proteins,intensity=colSums(data,na.rm=T))
meta$sample = rownames(meta)

meta$sample_origin = NA
meta$apicome = NA
meta$patient = NA
meta$zone = NA
meta$batch = NA
for(i in 1:nrow(meta)){
  s = str_split(rownames(meta)[i],"_")
  meta$apicome[i] = s[[1]][1]
  meta$zone[i] = s[[1]][2]
  batch = "1" 
  if (grepl("\\.",s[[1]][3])){
    sub_s = str_split(s[[1]][3],"\\.")
    patient = sub_s[[1]][1]
    batch = ifelse(sub_s[[1]][2]=="x","1","2")
  } else {
    patient = s[[1]][3]
  }
  meta$batch[i] = batch
  meta$patient[i] = patient
}

meta$sample_origin = paste(meta$apicome,meta$zone,sep="_")

meta$filter = F
meta$filter[meta$proteins < proteins_thresh] = T

ggplot(meta, aes(x=sample, weight=proteins,fill=filter)) + 
  geom_bar() +
  theme_bw() +
  xlab("") + ylab("Number of proteins identified") +
  scale_fill_manual(values=c("black","red")) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  geom_hline(yintercept=proteins_thresh, color="orange",linetype="dashed",linewidth=0.7) +
  coord_flip() + theme(legend.position="none")
ggsave(paste(dir_output,"/QC.pdf",sep=""),width=8,height=8,units="in")

```

```{r}
data[is.na(data)] = 0
```

# filter low quality samples
```{r}
write.csv(data,paste(dir_output,"all_samples.csv",sep="/"))
meta["basal_base_P13","filter"] = T # contaminated sample
data = data[,meta$filter == F]
```

## filter
```{r}
# data = data[data$filter == F,colnames(data)[colnames(data)!="non_nuc"]]
```
# create seurat object
```{r}
data[is.na(data)] = 0

rows = rownames(data)
data = data.frame(lapply(data, function(x) as.integer(x)))
rownames(data) = rows

seurat_obj = CreateSeuratObject(counts = data, meta.data=meta)

seurat_obj = NormalizeData(object = seurat_obj,normalization.method="RC",scale.factor = 1, verbose = F)
seurat_obj = FindVariableFeatures(object = seurat_obj)
seurat_obj = ScaleData(seurat_obj, features=rownames(seurat_obj), verbose = F)
seurat_obj = RunPCA(object = seurat_obj,features=VariableFeatures(object= seurat_obj),npcs=7, verbose = F)
```


# normilization and median of each condition
```{r}
data = matnorm(data)

for (cond in c("apical_tip","basal_tip","apical_base","basal_base")) {
  cond_samples = grepl(cond, colnames(data))
  num_samples = sum(cond_samples)
  if (num_samples == 0){
    next()
  } else if (num_samples == 1) {
      data[,cond] = data[,grepl(cond, colnames(data))]
      data[,paste(cond,"_SD",sep="")] = NA
  } else {
      data[,cond] = apply(data[,cond_samples],1,median, na.rm=T)
      data[,paste(cond,"_SD",sep="")] = apply(data[,cond_samples],1,sd,na.rm=T)     
  }
}
```

# apical/basal and tip/base score
```{r}
PN = 1e-6

data$gene = rownames(data)

# apical/basal in tip only, since no basal base:
data$a_b_ratio = log((data$apical_tip+PN) / (data$basal_tip+PN), 2)
```

# deseq2
```{r Warning=F}
EXP_THRESH = 0

Idents(seurat_obj) = seurat_obj$apicome 

s = subset(seurat_obj,subset=sample_origin %in% c("apical_tip","basal_tip")) # only tip

df = FindMarkers(s, ident.1=c("apical"),ident.2=c("basal"),test.use="DESeq2",
                    min.pct=0,min.diff.pct=0,min.cells.feature=0,min.cells.group=0,slot="counts",fc.name="log2_fc")
df[,c("pct.1","pct.2","log2_fc")] = NULL
colnames(df) = c("pval","qval")

```

## add the qval to the data
```{r Warning=F}
data_without_qvals = data
data = merge_quick(list(data,df))

data$expression_mean = (data$apical_tip+data$basal_tip)/2 

```

# MA
```{r}
QVAL_THRESH = 0.25

genes = c("ANPEP","SLC5A1","ITGA6","CASK","ACE2","GPA33")

plot = data
plot$signif = "no"
plot$signif[plot$qval < QVAL_THRESH] = "signif"
plot$signif[(plot$gene %in% genes) & (plot$signif == "signif")] = "selected"

plot$signif = factor(plot$signif,levels=c("no","signif","selected"))
plot = plot[order(plot$signif),]

plot$log10_expression = log(plot$expression_mean,10)
plot$log10_qval = -log(plot$qval,10)

ggplot(plot, aes(log10_expression,a_b_ratio,fill=signif)) +
  geom_point(aes(color=signif,size=signif),shape=21) +
  geom_text_repel(aes(label=gene,color=signif),data=plot[!plot$signif %in% c("no","signif"),],max.overlaps = 30,size=4) +
  xlab(expression("log10(max expression)")) + 
  ylab(expression("log2(apical/basal)")) +
  theme_bw(base_size=18)+
  scale_size_manual(values=c("no"=1,"signif"=1,"selected"=3))+
  scale_fill_manual(values=c("no"="black","signif"="blue","selected"="blue")) +
  scale_color_manual(values=c("no"="black","signif"="blue","selected"="red")) +
  theme(panel.grid.major=element_blank(),panel.grid.minor=element_blank(),legend.position="none")
ggsave(paste(dir_output,"/MA_proteins.pdf",sep=""),width=6,height=6,units="in")

```

```{r}
print(paste0("significant polarized proteins: ",nrow(drop_na(plot[plot$qval<=0.25,])),"   Which is ",(nrow(drop_na(plot[plot$qval<=0.25,])) / nrow(plot))*100," %"))
print(paste0("apical: ",nrow(drop_na(plot[(plot$qval<=0.25) & (plot$a_b_ratio>0),]))))
print(paste0("basal: ",nrow(drop_na(plot[(plot$qval<=0.25) & (plot$a_b_ratio<0),]))))
```

# Violin plot
## define groups of proteins, calculate pvalue, plot violins
```{r}
EXP_THRESH = 1e-5

plot = data[data$expression_mean >= EXP_THRESH,c("a_b_ratio","gene")]

control = "Other"
slcs = find_gene(plot,"^SLC",gene=T)
ribo = find_gene(plot,"^RP[LS]",gene=T)
mito = find_gene(plot,", mitochondrial")
histones = find_gene(plot,"histon")
matrisome = read.csv(paste0(directory,"/input/Matrisome_HomoS_Naba_et_al_2012.csv"))
matrisome = matrisome$Gene.Symbol

plot$group = "Other"
plot$group[plot$gene %in% matrisome] = "ECM"
plot$group[plot$gene %in% mito] = "Mitochondrial"
plot$group[plot$gene %in% ribo] = "Ribosomal"
plot$group[plot$gene %in% histones] = "Histones"
plot$group[plot$gene %in% slcs] = "SLC"

groups = c("Histones","ECM","Ribosomal","Mitochondrial","SLC")

plot$group = factor(plot$group,levels=c(groups,control))

comparisons = lapply(groups, function(group) c(control, group))
```

## calculate pvalues for groups
```{r}
p_values = data.frame(group1 = control,group2 = groups,pval = rep(NA,length(groups)), x = 1:length(groups))

for (group in groups){
  group1 = plot$a_b_ratio[plot$group==group]
  group2 = plot$a_b_ratio[plot$group!=group]
  p = wilcox.test(group1,group2,alternative="two.sided")
    # group1 = plot$a_b_ratio[plot$group==control]
  # group = plot$a_b_ratio[plot$group==group]
  # p = wilcox.test(group,alternative="two.sided")
  p_values$pval[p_values$group2==group] = p$p.value
}

p_values$qval = p.adjust(p_values$pval, method="BH")
```


## plot violin of all groups
```{r}
plot = drop_na(plot)
max_y = max(plot$a_b_ratio) * 1.2

plot = plot[plot$group!=control,]

median(plot$a_b_ratio[plot$group=="Ribosomal"],na.rm = T)

ggplot(data=plot, aes(x=group, y=a_b_ratio)) +
  geom_hline(yintercept=0,alpha=0.5,linetype="dashed") +
  geom_violin(alpha=1, fill="lightblue",trim=F) +
  geom_jitter(size=1, alpha=0.1, color="blue", position=position_jitter(width =.2)) +
  geom_boxplot(outlier.shape=NA, width=0.1) +
  theme_bw(base_size=18) + 
  theme(legend.position ="none",axis.text.x=element_text(color="black",angle=45,vjust =0.6),panel.grid = element_blank(),axis.text.y=element_text(color="black")) +
  xlab("") + ggtitle("") +
  ylab("log2(apical/basal) - Proteins") +
  lapply(1:nrow(p_values), function(i) {
    annotate("text",x=p_values$x[i],y=max_y,label=sprintf(paste0("p=",format.pval(p_values$qval[i],digits=2))), size=4)
  })
ggsave(paste(dir_output,"/groups_violin.pdf",sep=""),width=5,height=6,units="in")


```

# heatmap
## rearrange data
```{r}
h = data[,c("gene","qval","basal_tip","apical_tip")]
colnames(h) = c("gene","qval","Basal","Apical")
h$max = apply(h[,c("Basal","Apical")],1,max)
h$Apical = h$Apical/h$max
h$Basal = h$Basal/h$max
h$delta = h$Basal - h$Apical
h$gene = factor(h$gene, levels = h$gene[order(h$delta)]) # rearrange heatmap according to delta

h_long = pivot_longer(h[,c("gene","qval","Basal","Apical","delta")],-c(gene,qval,delta),names_to="apicome",values_to="rel_expression")
h_long$apicome = factor(h_long$apicome,levels=c("Basal","Apical"))

```

## plot function
```{r}
heatmap_prot = function(genes,title="",all=F){
  h_long_sub = h_long[h_long$gene %in% genes,]
  if (nrow(h_long_sub)==0){return()}
  if (all==T){size=4}else {size=18}
  p = ggplot(h_long_sub, aes(x = apicome, y = gene, fill = rel_expression)) +
    geom_tile(color = "black") +
    scale_fill_viridis() +
    theme_minimal(base_size=size) +
    theme(axis.text.x = element_text(color="black"),axis.text.y = element_text(color="black"),axis.ticks.x = element_blank(),axis.ticks.y = element_blank()) +
    xlab("") + ylab("") + ggtitle(title) +
    theme(legend.position="none",plot.title = element_text(size = 18)) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) 
  return(p)
}
```

## define pathways
```{r}
h_long$pathway = NA

mucus = c("ZG16","CLCA1","FCGBP","MUC3A","MUC13","MUC2","MUC6")
lipids = c('PNLIP','CD36','FABP2','FABP1','APOA4','APOA1','PCK1','MTTP','GK',
           'PCK2','APOB','PDK4','DGAT1','ACSL5','SLC27A4','DGAT2')
glucose = c('SLC5A1','SLC2A5','SLC2A2')
amino_acids = c('SLC36A1','SLC3A1','SLC6A19')
enzymes = c('ANPEP','SI','LCT','MGAM','ENPEP','DPP4','PEPD','LIPA','TMPRSS15')
xenobiotics = c('CYP3A4','CYP3A5','CYP2C9','CBR1','ABCC2','ABCG2','FMO5')
sulfur = data$gene[startsWith(data$gene,'SULT')]
ecm = find_gene(data,"^ITG",gene=T) 

h_long$pathway[h_long$gene %in% mucus] = "Mucus composition"
h_long$pathway[h_long$gene %in% lipids] = "Fatty-acid transport and processing"
h_long$pathway[h_long$gene %in% glucose] = "Monosacharide transport"
h_long$pathway[h_long$gene %in% amino_acids] = "Amino-acids transport"
h_long$pathway[h_long$gene %in% enzymes] = "Digestive enzymes" 
h_long$pathway[h_long$gene %in% xenobiotics] = "Xenobiotic metabolism"
h_long$pathway[h_long$gene %in% sulfur] = "Sulfotransferases"
h_long$pathway[h_long$gene %in% ecm] = "Integrins"

h_long_sub = h_long[!is.na(h_long$pathway),]

h_long_sub$pathway = factor(h_long_sub$pathway,
  levels=c("Amino-acids transport","Mucus composition", "Fatty-acid transport and processing", 
           "Integrins","Digestive enzymes","Xenobiotic metabolism","Sulfotransferases","Monosacharide transport"))

```

## plot heatmaps
```{r}
h_long_sub1 = h_long_sub[grepl("Digestive|Fatty|Amino|Integrin|Mucus|Xenobiotic|Sulfo|Monos",h_long_sub$pathway),]

ggplot(h_long_sub1,aes(y=apicome,x=gene,fill=rel_expression)) +
  geom_tile(color="black") +
  facet_grid(.~pathway, scales="free_x",space="free_x") +  
  scale_fill_viridis(name="Relative abundance",breaks = c(0, 1), labels = c("0", "1")) +
  theme_minimal(base_size=18) +
  theme(axis.text.x=element_text(color="black",angle=45,hjust=1,vjust =1.1),  
        axis.text.y=element_text(color="black"),
        panel.background=element_blank(),
        panel.grid=element_blank(),
        strip.background=element_blank(),
        axis.ticks=element_blank(),
        strip.text.x=element_text(angle = 0), 
        panel.spacing=unit(1, "lines"),
        legend.position = "right") +
  xlab("") + ylab("") + ggtitle("") 

ggsave(paste(dir_output,"/heatmap_long.pdf",sep=""),width=24,height=3,units="in")

```

# boxplots of figure 4A (Fig S4)
```{r}
h = data[,grepl("apical_tip_P|basal_tip_P",colnames(data))]
max_abundance = apply(h,1,max)
h = h / max_abundance
h$gene = rownames(h)
h_long = pivot_longer(h,-c(gene),names_to="sample",values_to="rel_expression")
h_long = separate(h_long,col="sample", into = c("apicome", "zone", "patient"), sep = "_")

h_long$zone = NULL
delta = h_long %>%
  group_by(gene, apicome) %>%
  summarise(mean_rel_expression = mean(rel_expression, na.rm = TRUE)) %>%
  spread(apicome, mean_rel_expression) %>%
  mutate(delta = apical - basal)

h_long = merge(h_long,delta[,c("gene","delta")],by="gene")

h_long$pathway = NA

h_long$pathway[h_long$gene %in% mucus] = "Mucus composition"
h_long$pathway[h_long$gene %in% lipids] = "Fatty-acid transport and processing"
h_long$pathway[h_long$gene %in% glucose] = "Monosacharide transport"
h_long$pathway[h_long$gene %in% amino_acids] = "Amino-acids transport"
h_long$pathway[h_long$gene %in% enzymes] = "Digestive enzymes" 
h_long$pathway[h_long$gene %in% xenobiotics] = "Xenobiotic metabolism"
h_long$pathway[h_long$gene %in% sulfur] = "Sulfotransferases"
h_long$pathway[h_long$gene %in% ecm] = "Integrins"

h_long_sub = h_long[!is.na(h_long$pathway),]

h_long_sub$pathway = factor(h_long_sub$pathway,
  levels=c("Amino-acids transport","Mucus composition", "Fatty-acid transport and processing", 
           "Integrins","Digestive enzymes","Xenobiotic metabolism","Sulfotransferases","Monosacharide transport"))
h_long_sub$apicome = str_to_title(h_long_sub$apicome)
h_long_sub$apicome = factor(h_long_sub$apicome,levels=c("Basal","Apical"))
```

```{r}
create_violin = function(pathway,legend=F,ylabel=T) {
  to_plot = h_long_sub[h_long_sub$pathway==pathway,]
  to_plot$gene = factor(to_plot$gene,levels=unique(to_plot$gene[order(to_plot$delta,decreasing=T)]))
  p = ggplot(to_plot, aes(x = apicome, y = rel_expression)) +
    facet_grid(~gene) +
    geom_boxplot(aes(fill=apicome),outlier.shape = NA) +
    theme_bw() + 
    # geom_jitter(size=0.5, alpha=0.7, position=position_jitter(width=0.2)) +
    scale_fill_manual(values=c("orange","lightblue"))+
    theme(panel.grid = element_blank(), 
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text = element_text(color = "black")) +
    xlab(NULL) + ggtitle(pathway) +
    ylab("") +
    scale_y_continuous(breaks = c(0, 1), labels = c(0, 1)) 
  if (legend==T){
    p = p + theme(legend.position="right",
        legend.spacing.y = unit(0, "cm"),
        legend.background=element_rect(colour="black",linewidth=0.5),
        legend.direction = "vertical",
        legend.text = element_text(size = 14),  
        legend.key.size = unit(1.5, "lines"),
        legend.title = element_blank())
  } else {p = p + theme(legend.position = "none")}
  if (ylabel==T){p = p + ylab("Relative abundance")}
  return (p)
}

plots = list()
plots[[1]] = create_violin("Xenobiotic metabolism", legend=T) 
plots[[2]] = create_violin("Mucus composition")
plots[[3]] = create_violin("Digestive enzymes")
plots[[4]] = create_violin("Fatty-acid transport and processing")
plots[[5]] = create_violin("Monosacharide transport") + create_violin("Integrins",ylabel=F)
plots[[6]] = create_violin("Amino-acids transport") + create_violin("Sulfotransferases",ylabel=F)

legend = get_legend(plots[[1]])
plots[[1]] = plots[[1]] + theme(legend.position="none")
first_plot_with_legend = plot_grid(plots[[1]], legend, ncol = 2, rel_widths = c(4, 1))

combined_plot = plot_grid(first_plot_with_legend, 
                          plots[[2]], plots[[3]], plots[[4]], plots[[5]], plots[[6]],
                          ncol=1, rel_heights=rep(1,length(plots))) +
  theme(plot.margin = unit(c(0.07, 0.07, 0.07, 0.07), "cm"))

ggsave(paste(dir_output,"/boxplots.pdf",sep=""),width=10.5,height=12,units="in")
# shell.exec(paste(dir_output,"/boxplots.pdf",sep=""))
```


# pathview kegg pathway visualization
```{r}
translation_human = read.csv(paste(root_dir,"human_proteomics","input","human_ensemble_conversion.csv",sep="/"))
translation_human = translation_human[!duplicated(translation_human$external_gene_name),]

kegg_draw_pathway(gene_names=data$gene,values=data$a_b_ratio,pathway.id="04975",filename="FattyAcid",limit =1)
kegg_draw_pathway(gene_names=data$gene,values=data$a_b_ratio,pathway.id="04973",filename="Carbohydrates",limit =1)
```

# filter non-nuclear only proteins
## find non nuclear genes
```{r}
subcellular1 = read.csv(paste0(directory,"/input/uniprot_GO_human.csv"))
subcellular2 = read.csv(paste0(directory,"/input/HPA_subcellular_location_v23.0.csv"),row.names = 1)

subcellular_long = subcellular1[,c("Gene_name","Reviewed")] %>%
  mutate(Full_names = Gene_name) %>%
  separate_rows(Gene_name, sep = " ")
subcellular_long$Reviewed = NULL

subcellular = merge(subcellular2[,c("Gene.name","GO.id")],subcellular_long,by.x="Gene.name",by.y="Gene_name",all.x=T,all.y=F)
subcellular = subcellular[, c("Full_names", "GO.id")]

subcellular$first_gene = sapply(str_split(subcellular$Full_names, " "), function(x) x[1])
subcellular$gene_name_len = nchar(subcellular$Full_names)

subcellular = subcellular %>%
  group_by(first_gene) %>%
  filter(gene_name_len == max(gene_name_len)) %>%
  ungroup()

subcellular = distinct(subcellular[,c("Full_names","GO.id")])
```

```{r}
unique_values = subcellular %>%
  pull(GO.id) %>%  # Extract the column as a vector
  str_split(";") %>%     
  unlist() %>%           # Flatten the list into a vector
  trimws() %>%           # Trim any leading or trailing whitespace
  unique()               # Get unique values

non_nuclear_id = unique_values[!grepl("Nuc",unique_values,ignore.case = T)]
non_nuclear_id = sapply(str_split(non_nuclear_id, "\\(GO:"), function(x) x[2])
non_nuclear_id = non_nuclear_id[!is.na(non_nuclear_id)]
non_nuclear_id = gsub("\\)","",non_nuclear_id)
subcellular$nuc = T
for (id in non_nuclear_id){
  cur_grepl = grepl(id, subcellular$GO.id)
  subcellular$nuc[cur_grepl] = F
}


nuclear_genes = subcellular$Full_names[subcellular$nuc==T]
data$filter = sapply(rownames(data), function(gene) any(grepl(paste0("(^|\\s)",gene, "($|\\s)"), nuclear_genes)))
data$filter[grepl("^HIST|^H2|^H1|^H3",rownames(data))] = T # remove histones

```

## filter matrisome
```{r}
data$filter[rownames(data) %in% matrisome] = T
```

## check correlation apical/basal between filtered and non-filtered data
```{r}
data_cyto = data[data$filter==F,]
data_cyto = matnorm(data_cyto)
data_cyto = data_cyto[,grepl("_P",colnames(data_cyto))]

pn = min(data_cyto[data_cyto>0])
pn = 0
data_cyto$expression_mean = apply(data_cyto,1,mean,na.rm=T)
data_cyto$apical = apply(data_cyto[,grepl("apical_tip_",colnames(data_cyto))],1,median)
data_cyto$basal = apply(data_cyto[,grepl("basal_tip_",colnames(data_cyto))],1,median)
data_cyto$log2_fc = log2((data_cyto$apical + pn) / (data_cyto$basal + pn))
data_cyto$gene = rownames(data_cyto)

data_all = data[,grepl("_P",colnames(data))]
pn = min(data_all[data_all>0])
pn = 0
data_all$expression_mean = apply(data_all,1,mean,na.rm=T)
data_all$apical = apply(data_all[,grepl("apical_tip_",colnames(data_all))],1,median)
data_all$basal = apply(data_all[,grepl("basal_tip_",colnames(data_all))],1,median)
data_all$log2_fc = log2((data_all$apical + pn) / (data_all$basal + pn))
data_all$gene = rownames(data_all)

columns = c("gene","expression_mean","log2_fc","apical","basal")
data_merged = merge(data_cyto[,columns],data_all[,columns],by="gene",suffixes=c("_cyto","_all"),all=F)

cor(data_merged$log2_fc_all, data_merged$log2_fc_cyto, method = "spearman")

plot = data_merged
ggplot(plot, aes(log2_fc_all, log2_fc_cyto)) +
  geom_point() +
  stat_cor(method="spearman",digits=2) +
  theme_bw(base_size=14) +
  xlab("log2(apical/basal) - All proteins") +  ylab("log2(apical/basal) - Cytoplasmic only") +
  geom_hline(yintercept=0, color="black",alpha = 0.2,linetype="dashed") +
  geom_vline(xintercept=0, color="black",alpha = 0.2,linetype="dashed") +
  theme(panel.grid=element_blank(),
        axis.text=element_text(color="black"),
        legend.position = "none") +
  ggtitle("")
ggsave(paste(dir_output,"/filtering_correlation.pdf",sep=""),width=5,height=5,units="in")
```

# heatmaps on filtered data
```{r}
h = data_cyto[,c("gene","basal","apical")]
colnames(h) = c("gene","Basal","Apical")
h$max = apply(h[,c("Basal","Apical")],1,max)
h$Apical = h$Apical/h$max
h$Basal = h$Basal/h$max
h$delta = h$Basal - h$Apical
h$gene = factor(h$gene, levels = h$gene[order(h$delta)])

h_long = pivot_longer(h[,c("gene","Basal","Apical","delta")],-c(gene,delta),names_to="apicome",values_to="rel_expression")
h_long$apicome = factor(h_long$apicome,levels=c("Basal","Apical"))

```

```{r}
h_long$pathway = NA

h_long$pathway[h_long$gene %in% mucus] = "Mucus composition"
h_long$pathway[h_long$gene %in% lipids] = "Fatty-acid transport and processing"
h_long$pathway[h_long$gene %in% glucose] = "Monosacharide transport"
h_long$pathway[h_long$gene %in% amino_acids] = "Amino-acids transport"
h_long$pathway[h_long$gene %in% enzymes] = "Digestive enzymes" 
h_long$pathway[h_long$gene %in% xenobiotics] = "Xenobiotic metabolism"
h_long$pathway[h_long$gene %in% sulfur] = "Sulfotransferases"
h_long$pathway[h_long$gene %in% ecm] = "Integrins"

h_long_sub = h_long[!is.na(h_long$pathway),]
h_long_sub$pathway = factor(h_long_sub$pathway,
  levels=c("Amino-acids transport","Mucus composition", "Fatty-acid transport and processing", 
           "Integrins","Digestive enzymes","Xenobiotic metabolism","Sulfotransferases","Monosacharide transport"))

```

```{r}
h_long_sub1 = h_long_sub[grepl("Digestive|Fatty|Amino|Integrin|Mucus|Xenobiotic|Sulfo|Monos",h_long_sub$pathway),]

ggplot(h_long_sub1,aes(y=apicome,x=gene,fill=rel_expression)) +
  geom_tile(color="black") +
  facet_grid(.~pathway, scales="free_x",space="free_x") +  
  scale_fill_viridis(name="Relative abundance",breaks = c(0, 1), labels = c("0", "1")) +
  theme_minimal(base_size=18) +
  theme(axis.text.x=element_text(color="black",angle=45,hjust=1,vjust =1.1),  
        axis.text.y=element_text(color="black"),
        panel.background=element_blank(),
        panel.grid=element_blank(),
        strip.background=element_blank(),
        axis.ticks=element_blank(),
        strip.text.x=element_text(angle = 0), 
        panel.spacing=unit(1, "lines"),
        legend.position = "right") +
  xlab("") + ylab("") + ggtitle("") 

ggsave(paste(dir_output,"/filtered_heatmap_long.pdf",sep=""),width=24,height=3,units="in")
```

# export data
```{r}
write.csv(meta,paste(dir_output,"samples_metadata.csv",sep="/"))
write.csv(data,paste(dir_output,"processed_data.csv",sep="/"), row.names=F)
saveRDS(seurat_obj, paste(dir_output,"data.rds",sep="/"))
```
